---
title: 'Reproducible Research: Reproducible Data Peer Assessment 1'
author: "Tai Chee Ming"
date: "18 September 2017"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```

#Code for reading in the dataset and/or processing the data

Set up working directory
```{r, echo=TRUE}
setwd("D:/Downloads/Coursera/Reproducible Research/Course Project 1A")
```

URL to download the data zip file
```{r, echo=TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```


Download the zip file into workspace
```{r, echo=TRUE}
download.file(fileUrl, destfile="activity.zip")
```

Unzip the file, read the file using read.csv() to load the data into R
```{r, echo=TRUE}
data <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, sep = ",", na.strings = "NA")
```

Look at the summary of the data
```{r, echo=TRUE}
summary(data)
```

Look at the structure of the data
```{r, echo=TRUE}
str(data)
```

Looking at the first 6 rows of the dataset:
```{r, echo=TRUE}
head(data)
```

Converting the "date" variable to a Date class and the "interval" variable to a factor
```{r, echo=TRUE}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- factor(data$interval)
```

What is mean total number of steps taken per day?
Subsetting the dataset to ignore missing values
```{r, echo=TRUE}
NA_index <- is.na(as.character(data$steps))
data_no_NA <- data[!NA_index,]
head(data_no_NA)
```

Aggregating the number of steps taken each day
Creating a data frame with the steps taken for each day
```{r, echo=TRUE}
steps_each_day <- aggregate(steps ~ date, data = data_no_NA, sum)
```

Adding column names to the created data frame
```{r, echo=TRUE}
colnames(steps_each_day) <- c("date", "steps")
```

#Histogram of the total number of steps taken each day.
```{r, echo=TRUE}
hist(as.numeric(steps_each_day$steps), breaks = 20, col = "Blue", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
```

#Mean and median number of steps taken each day
```{r, echo=TRUE}
mean(steps_each_day$steps)
median(steps_each_day$steps)
```

What is the average daily activity pattern?
Calculating the average number of steps taken, averaged across all days.
```{r, echo=TRUE}
steps_per_interval <- aggregate(data_no_NA$steps, by=list(interval=data_no_NA$interval), FUN=mean)
```

Adding columns names
```{r, echo=TRUE}
colnames(steps_per_interval) <- c("interval", "average_steps")
```

#Time series plot of the average number of steps taken
```{r, echo=TRUE}
plot(as.integer(levels(steps_per_interval$interval)), steps_per_interval$average_steps, type="l",xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="blue")
```

The maximum number of average steps
```{r, echo=TRUE}
max_steps <- max(steps_per_interval$average_steps)
max_steps
```

#The 5-minute interval that, on average, contains the maximum number of steps
```{r, echo=TRUE}
intervale_max_steps<-steps_per_interval[which.max(steps_per_interval$average_steps),]$interval
intervale_max_steps
```

##The total number of missing values in the dataset (for each variable)

For the "steps" variable:
```{r, echo=TRUE}
sum(is.na(as.character(data$steps)))
```

For the "date" variable:
```{r, echo=TRUE}
sum(is.na(as.character(data$date)))
```

For the "interval" variable:
```{r, echo=TRUE}
sum(is.na(as.character(data$interval)))
```

#Code to describe and show a strategy for imputing missing data
Finding the indices of missing values (NAs)
```{r, echo=TRUE}
NA_index <- which(is.na(as.character(data$steps)))
complete_data <- data
```
Imputing missing values using the mean for that 5-minute interval
```{r, echo=TRUE}
complete_data[NA_index, ]$steps<-unlist(lapply(NA_index, FUN=function(NA_index){
        steps_per_interval[data[NA_index,]$interval==steps_per_interval$interval,]$average_steps
}))
```
Checking the complete data with the summary and str methods
```{r, echo=TRUE}
summary(complete_data)
str(complete_data)
```

Making a histogram of the total number of steps taken each day for the complete dataset.
Creating a data frame with the steps taken for each day
```{r, echo=TRUE}
steps_each_day_complete <- aggregate(steps ~ date, data = complete_data, sum)
```
Adding column names to the created data frame
```{r, echo=TRUE}
colnames(steps_each_day_complete) <- c("date", "steps")
```

#Histogram of the total number of steps taken each day after missing values are imputed
```{r, echo=TRUE}
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
```

Calculating the mean and median total number of steps taken per day for the complete dataset.
```{r, echo=TRUE}
mean(steps_each_day_complete$steps)
median(steps_each_day_complete$steps)
```

Are there differences in activity patterns between weekdays and weekends?
Creating a factor variable "day "to store the day of the week:
```{r, echo=TRUE}
complete_data$day <- as.factor(weekdays(complete_data$date))
```

Creating a logical variable "is_weekday" (weekday=TRUE, weekend = FALE) :
```{r, echo=TRUE}
complete_data$is_weekday <- ifelse(!(complete_data$day %in% c("Saturday","Sunday")), TRUE, FALSE) 
```


Calculating the average number of steps for weekdays
```{r, echo=TRUE}
weekdays_data <- complete_data[complete_data$is_weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)
```


Calculating the average number of steps for weekends
```{r, echo=TRUE}
weekends_data <- complete_data[!complete_data$is_weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)
```

Adding columns names
```{r, echo=TRUE}
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")
```

Adding a column to indecate the day
```{r, echo=TRUE}
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"
```

Merging the two togather
```{r, echo=TRUE}
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)
```

Converting the day variabke to a factor
```{r, echo=TRUE}
week_data$day <- as.factor(week_data$day)
```

#Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends.
```{r, echo=TRUE}
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, layout = c(1,2), type ="l", ylab="Number of Steps")
```

The plot shows that that activity on the weekends tends to be more spread out over the day compared to the weekdays.
